
# Getting started


## Checklist (before the workshop)

Install the following software in advance in order to avoid
unnecessary delays and leaving more time for the workshop contents.


* [R (version >4.1.0)](https://www.r-project.org/) 

* [RStudio](https://www.rstudio.com/products/rstudio/download/);
  choose "Rstudio Desktop" to download the latest version. Optional
  but preferred. For further details, check the [Rstudio home
  page](https://www.rstudio.com/).
  
* For Windows users: [Rtools](https://cran.r-project.org/bin/windows/Rtools/rtools40.html);
  Follow the instructions to install the toolkit. This might be required to compile some of the 
  packages required on this course.

* Install and load the required R packages


## Support and resources


For online support on installation and other matters, you can join us at:

 * Users: [miaverse Gitter channel](https://gitter.im/microbiome/miaverse?utm_source=badge&utm_medium=badge&utm_campaign=pr-badge&utm_content=badge)
 * Developers: [Bioconductor Slack](https://bioc-community.herokuapp.com) #microbiomeexperiment channel (ask for an invitation)



## Installing and loading the required R packages

This section shows how to install and load all required packages into
the R session. Only uninstalled packages are installed.

```{r warning = FALSE, message = FALSE}

if (!requireNamespace("BiocManager")) {
  install.packages("BiocManager")
}

# List of packages that we need from cran and bioc 
packages <- c("ggplot2", "pheatmap", "stringr", "igraph", "ANCOMBC",
             "microbiome", "httpuv", "microbiomeDataSets", "mia", "caret", "ranger",
            "dplyr", "miaViz", "knitr", "kableExtra", "vegan", "ecodist", "biclust",
            "patchwork", "pdp", "MLmetrics", "precrec", "MOFA2")

```
 
The following script tries to load all required packages, and if they are not available, installs them.
```{r warning = FALSE, message = FALSE}
new.pkgs <- packages[!(packages %in% installed.packages()[, "Package"])]

if (length(new.pkgs)) {
    BiocManager::install(new.pkgs, ask=T)
}

sapply(packages, require, character.only = TRUE)
```
