```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(mia)
library(miaViz)
library(dplyr)
library(stringr)

mae <- microbiomeDataSets::HintikkaXOData()

# Drop off those bacteria that do not include information in Phylum or lower levels
mae[[1]] <- mae[[1]][!is.na(rowData(mae[[1]])$Phylum), ]

# Clean taxonomy data, so that names do not include addtional characters
rowData(mae[[1]]) <- DataFrame(apply(rowData(mae[[1]]), 2, 
                                     str_remove, pattern = "._[0-9]__"))

# For simplicity, classify all high-fat diets as high-fat, and all the low-fat 
# diets as low-fat diets
colData(mae)$Diet <- ifelse(colData(mae)$Diet == "High-fat" | 
                              colData(mae)$Diet == "High-fat + XOS", 
                            "High-fat", "Low-fat")

# Calculates relative abundances, and stores the table to assays
mae[[1]] <- transformCounts(mae[[1]], method = "relabundance")

# agglomerate the microbiota experiment to Phylum
se_phylum <- agglomerateByRank(mae[[1]], rank = "Phylum")
```


# Beta diversity

Beta diversity is another name for sample dissimilarity. It quantifies
differences in the overall taxonomic composition between two samples.

Common indices include Bray-Curtis, Unifrac, Jaccard index, and the
Aitchison distance. For more background information
and examples, you can check the dedicated section in [online
book](https://microbiome.github.io/OMA/microbiome-diversity.html#beta-diversity).


## Examples of PCoA with different settings

Beta diversity estimation generates a (dis)similarity matrix that
contains for each sample (rows) the dissimilarity to any other sample
(columns).

This complex set of pairwise relations can be visualized in
informative ways, and even coupled with other explanatory
variables. As a first step, we compress the information to a lower
dimensionality, or fewer principal components, and then visualize
sample similarity based on that using ordination techniques, such as
Principal Coordinate Analysis (PCoA). PCoA is a non-linear dimension
reduction technique, and with Euclidean distances it is is identical
to the linear PCA (except for potential scaling).

We typically retain just the two (or three) most informative top
components, and ignore the other information. Each sample has a score
on each of these components, and each component measures the variation
across a set of correlated taxa. The top components are then easily
visualized on a two (or three) dimensional display.

Let us next look at some concrete examples.


### PCoA for ASV-level data with Bray-Curtis

Let us start with PCoA based on a Bray-Curtis dissimilarity matrix
calculated at Genus level abundances.


```{r pcoa_asv_bc, message=FALSE}
# Pick the relative abundance table
rel_abund_assay <- assays(mae[[1]])$relabundance

# Calculates Bray-Curtis distances between samples. Because taxa is in
# columns, it is used to compare different samples. We transpose the
# assay to get taxa to columns
bray_curtis_dist <- vegan::vegdist(t(rel_abund_assay), method = "bray")

# PCoA
bray_curtis_pcoa <- ecodist::pco(bray_curtis_dist)

# All components could be found here: 
# bray_curtis_pcoa$vectors
# But we only need the first two to demonstrate what we can do:
bray_curtis_pcoa_df <- data.frame(pcoa1 = bray_curtis_pcoa$vectors[,1], 
                                  pcoa2 = bray_curtis_pcoa$vectors[,2])

# Create a plot
bray_curtis_plot <- ggplot(data = bray_curtis_pcoa_df, aes(x=pcoa1, y=pcoa2)) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2", 
       title = "Bray-Curtis PCoA") +
  theme(title = element_text(size = 10)) # makes titles smaller

bray_curtis_plot
```

## Highlighting external variables 

We can map other variables on the same plot for example by coloring
the points accordingly.

The following is an example with a discrete grouping variable (Diet) shown with colors:

```{r pcoa_genus}
# Add diet information to data.frame
bray_curtis_pcoa_df$Diet <- colData(mae)$Diet
# Creates a plot
plot <- ggplot(data = bray_curtis_pcoa_df, aes_string(x = "pcoa1", y = "pcoa2", color = "Diet")) +
  geom_point() +
  labs(title = paste0("PCoA beta diversity ordination for microbiome samples"), x = paste0("PC1 (p = ", p_values[["pcoa1"]], ")"), y = paste0("PC2 (p = ", p_values[["pcoa2"]], ")")) +
  theme_bw(12) 

plot
```

PCoA plot in some cases could also be overlayed with a continuous variable. ([see example](https://microbiome.github.io/course_2022_miaverse/beta-diversity.html#pcoa-plot-with-continuous-variable))


## Estimating associations with an external variable

Next to visualizing whether any variable is associated with
differences between samples, we can also quantify the strength of the
association between community composition (beta diversity) and
external factors.

The standard way to do this is to perform a so-called permutational
multivariate analysis of variance (PERMANOVA). This method takes as
input the abundance table, which measure of distance you want to base
the test on and a formula that tells the model how you think the
variables are associated with each other. 

```{r permanova}
# First we get the relative abundance table
rel_abund_assay <- assays(mae[[1]])$relabundance

# again transpose it to get taxa to columns
rel_abund_assay <- t(rel_abund_assay)

# then we can perform the method
permanova_diet <- vegan::adonis(rel_abund_assay ~ Diet,
                                  data = colData(mae),
                                  permutations = 99)

# we can obtain a the p value for our predictor:
print(paste0("The test result p-value: ", 
              as.data.frame(permanova_diet$aov.tab)["Diet", "Pr(>F)"]))

```  

The diet variable is significantly associated with
microbiota composition (p-value is less than 0.05).

We can visualize those taxa whose abundances drive the
differences between diets. We first need to extract the model
coefficients of taxa:


```{r permanova_coefs}
# Gets the coefficients
coef <- coefficients(permanova_diet)["Diet1",]

# Gets the highest coefficients
top.coef <- sort(head(coef[rev(order(abs(coef)))],20))

# Plots the coefficients
top_taxa_coeffient_plot <- ggplot(data.frame(x = top.coef,
                                             y = factor(names(top.coef),
					     unique(names(top.coef)))),
                                  aes(x = x, y = y)) +
  geom_bar(stat="identity") +
  labs(x="", y="", title="Top Taxa") +
  theme_bw()

top_taxa_coeffient_plot
```

The above plot shows taxa as code names, and it is hard to tell which
bacterial groups they represent. However, it is easy to add human readable
names. We can fetch those from our rowData. Here we use Genus level names:

```{r}
# Gets corresponding Genus level names and stores them to top.coef
names <- rowData(mae[[1]])[names(top.coef), ][,"Genus"]

# Adds new labels to the plot
top_taxa_coeffient_plot <- top_taxa_coeffient_plot +
  scale_y_discrete(labels = names) # Adds new labels
top_taxa_coeffient_plot
```

The same test can be conducted using the oridination from PCoA as follows:

```{r}
bray_curtis_pcoa_df$Diet <- colData(mae)$Diet
p_values <- list()
for(pc in c("pcoa1", "pcoa2")){
  # Creates a formula from objects
  formula <- as.formula(paste0(pc, " ~ ", "Diet"))
  # Does the permanova analysis
  p_values[[pc]] <- vegan::adonis(formula, data = bray_curtis_pcoa_df,
                                  permutations = 9999, method = "euclidean"
  )$aov.tab["Diet", "Pr(>F)"]
}

# Creates a plot
plot <- ggplot(data = bray_curtis_pcoa_df, aes_string(x = "pcoa1", y = "pcoa2", color = "Diet")) +
  geom_point(size = 3) +
  labs(title = paste0("PCoA beta diversity ordination for microbiome samples"), x = paste0("PC1 (p = ", p_values[["pcoa1"]], ")"), y = paste0("PC2 (p = ", p_values[["pcoa2"]], ")")) +
  theme_bw(12) 

plot
```

There are many alternative and complementary methods for analysing
community composition. For more examples, see a dedicated section on
beta diversity in the [online
book](https://microbiome.github.io/OMA/microbiome-diversity.html#beta-diversity).

## Community typing


A dedicated section presenting examples on community typing is in the
[online book](https://microbiome.github.io/OMA/microbiome-community.html#community-typing).
