```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(mia)
library(miaViz)
library(stringr)

mae <- microbiomeDataSets::HintikkaXOData()

# Drop off those bacteria that do not include information in Phylum or lower levels
mae[[1]] <- mae[[1]][!is.na(rowData(mae[[1]])$Phylum), ]

# Clean taxonomy data, so that names do not include addtional characters
rowData(mae[[1]]) <- DataFrame(apply(rowData(mae[[1]]), 2, 
                                     str_remove, pattern = "._[0-9]__"))

# For simplicity, classify all high-fat diets as high-fat, and all the low-fat 
# diets as low-fat diets
colData(mae)$Diet <- ifelse(colData(mae)$Diet == "High-fat" | 
                              colData(mae)$Diet == "High-fat + XOS", 
                            "High-fat", "Low-fat")

# Calculates relative abundances, and stores the table to assays
mae[[1]] <- transformCounts(mae[[1]], method = "relabundance")
```

# Unsupervised learning

Unsupervised learning is a part of machine learning where we try to find information
from unknown data. It is also called data mining. Usually this means finding of
clusters, for instance. Cluster refers to group of samples/features that are similar
between each other. For example, based on clinical data we can try to find patient 
groups that have similar response to used drug.

## Biclustering

Biclustering is a clustering method, which simultaneously clusters rows and columns.
In this example, the aim is to find clusters where subset of taxa share similar
pattern over subset of metabolites. In our case, we try to find clusters where 
taxa and metabolites correlate similarly. 

Check more from OMA which has dedicated chapter on 
[biclustering](https://microbiome.github.io/OMA/biclustering.html). 

First, we subset metabolite data so that it includes fewer metabolites. We use 
coefficient of variation as a subset criteria. Meaning of this is that, if concentration
of metabolite does not vary, concentration does not differ between samples. 

By doing that, we avoid unnecessary multiple testing, and we focus only on interesting
metabolites. 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
# Threshold: metabolites whose (cv > +threshold or cv < -threshold), will be included
cv_threshold <- 0.5
metabolite_trans <- "nmr"

# Get the data
metabolite_tse <- mae[[2]]

# Calculate coefficient of variation of individual metabolites
df <- data.frame(cv = apply(assay(metabolite_tse, metabolite_trans), 1, 
                            function(x){sd(x)/mean(x)}))

# Plot them as a histogram, and show a line that is used as a threshold
plot <- ggplot(df, aes(x = cv)) +
  geom_histogram(bins = 50, color="darkred", fill="lightblue") +
  labs(x = "CV", y = "metabolite frequency", 
       title = "Distribution of coefficient of 
       variation of log10 concentration of metabolites") +
  geom_vline(xintercept = cv_threshold, color = "red") +
  geom_text(aes(cv_threshold, 6, label = 
                  paste0("CV threshold (", cv_threshold, ")"), vjust = 2, angle=90)) +
  geom_vline(xintercept = -cv_threshold, color = "red") +
  geom_text(aes(-cv_threshold, 6, label = 
                  paste0("CV threshold (", -cv_threshold, ")"), vjust = -1, angle=90))

plot
```

Subsetting metabolomic data:

```{r, warning=FALSE, message=FALSE}
# Get those metabolites that are over threshold
metabolites_over_th <- rownames(df[abs(df$cv) > cv_threshold, , drop = FALSE])
# Ignore those metabolites that do not have name / are NA
metabolites_over_th <- metabolites_over_th[!str_detect(metabolites_over_th, "NA")]
```

After that, we subset microbiome data by filtering rarest taxa off-

```{r, warning=FALSE, message=FALSE}
rank <- "Genus"
prevalence <- 0.2
detection <- 0.001
taxa_trans <-  "clr"

# Get bacterial data
taxa_tse <- mae[[1]]
# Agglomerate at Genus level
taxa_tse <- agglomerateByRank(taxa_tse, rank = rank)
# Do CLR transformation
taxa_tse <- transformSamples(taxa_tse, method = "clr", pseudocount = 1)

# Subset metabolite data
metabolite_tse <- metabolite_tse[metabolites_over_th, ]

# Subset bacterial data by its prevalence. Bacteria whose prevalences are over 
# threshold are included
taxa_tse <- subsetByPrevalentTaxa(taxa_tse, 
                                  prevalence = prevalence, 
                                  detection = detection)

# Remove uncultured and ambiguous(as it's hard to interpret their results)
taxa_tse <- taxa_tse[-grep("uncultured|Ambiguous_taxa", names(taxa_tse)),]
```

After subsetting, we can perform cross-correlation analysis. With cross-correlation
analysis we can answer the question "If the abundance of this taxon is high, is the 
concentration of metabolite high?", for instance.

Check OMA book's dedicated chapther on 
[multi-assay analyses](https://microbiome.github.io/OMA/multi-assay_analyses.html).  

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=8}
library(pheatmap)

# Cross correlate data sets
correlations <- testExperimentCrossCorrelation(taxa_tse, metabolite_tse, 
                                               abund_values1 = "clr", abund_values2 = "nmr",
                                               method = "spearman", mode = "matrix")

# For plotting purpose, convert p-values, under 0.05 are marked with "X"
p_threshold <- 0.01
p_values <- ifelse(correlations$p_adj<p_threshold, "X", "")

# Scale colors
breaks <- seq(-ceiling(max(abs(correlations$cor))), ceiling(max(abs(correlations$cor))), 
              length.out = ifelse( max(abs(correlations$cor))>5, 
                                   2*ceiling(max(abs(correlations$cor))), 10 ) )
colors <- colorRampPalette(c("darkblue", "blue", "white", 
                             "red", "darkred"))(length(breaks)-1)

# Create a heatmap
pheatmap(correlations$cor, display_numbers = p_values,
               main = paste0("Correlations between bacteria and metabolites 
              (statistically significant associations (p < 0.05) marked with X)"),
               fontsize = 10,
               breaks = breaks,
               color = colors, 
               fontsize_number = 10)
```

Finally, we can find biclusters from cross-correlation data.

```{r}
# Load package
library(biclust)

# Find biclusters
bc <- biclust(correlations$cor, method=BCPlaid(), fit.model = y ~ m,
              background = TRUE, shuffle = 100, back.fit = 0, max.layers = 10,
              iter.startup = 10, iter.layer = 100, verbose = FALSE)

bc
```

Now, we can add bicluster information into the heatmap that we already made.

```{r, message=FALSE, warning=FALSE, fig.height=10, fig.width=8}
# Convert boolean values into factors
bicluster_columns <- data.frame(apply(bicluster_columns, 2, as.factor))
bicluster_rows <- data.frame(apply(bicluster_rows, 2, as.factor))

# Adjust colors for all clusters
if( ncol(bicluster_rows) > ncol(bicluster_columns) ){
  cluster_names <- colnames(bicluster_rows)
} else {
  cluster_names <- colnames(bicluster_columns)
}
annotation_colors <- list()
for(name in cluster_names){
  annotation_colors[[name]] <- c("TRUE" = "red", "FALSE" = "white")
}

# Get correlation values that are over thresholds
p_threshold <- 0.01
corr_threshold <- 0.6
corr_values <- ifelse(correlations$p_adj<p_threshold & 
                          abs(correlations$cor)>corr_threshold , round(correlations$cor,1), "")

# Create a heatmap
pheatmap(correlations$cor,
         annotation_col = bicluster_columns, 
         annotation_row = bicluster_rows,
         annotation_colors = annotation_colors,
         display_numbers = corr_values,
               main = paste0("Correlations between bacteria and metabolites 
              (correlation over threshold (corr > ", corr_threshold,
              ", p < ", p_threshold,") marked)"),
               fontsize = 10,
               breaks = breaks,
               color = colors, 
               fontsize_number = 6, 
         number_color = "yellow",
         annotation_legend = FALSE)
```
